---
title: "aeolus vs openair — numerical parity"
subtitle: "How aeolus metrics compare to the R openair reference implementation"
format:
  html:
    toc: true
    code-fold: show
    theme: cosmo
    embed-resources: true
execute:
  echo: true
  warning: false
---

## Why this document exists

`aeolus` aims to be a Python-native replacement for core openair workflows.
For users to trust that, "works like openair" can't just be a design claim —
it has to be measurable. This page shows the evidence.

Every metric on this page is computed two ways on the same input data:

1. **R openair** computed the reference values on the pinned version shown
   in _Provenance_ below. Outputs are checked into the aeolus repo at
   `tests/fixtures/openair/`.
2. **aeolus** computes the same metrics from the same raw data.

The two are then compared value-by-value. A failing comparison fails CI,
and this document will not render until parity is restored.

::: {.callout-note}
The R code on this page is shown for documentation — it isn't executed
when the page renders, so you don't need R or openair installed to read
this. See `scripts/validation/generate_openair_fixtures.R` for the exact
commands that produced the reference fixtures.
:::

## Provenance

```{python}
#| echo: false
#| label: provenance
import pandas as pd
from pathlib import Path

FIXTURES = Path("../../tests/fixtures/openair")
prov = pd.read_csv(FIXTURES / "provenance.csv")
prov.T.rename(columns={0: "value"}).style.set_caption("Fixture provenance")
```

## Dataset

**Site**: London Marylebone Road (MY1), AURN.

**Period**: calendar year 2023 (8,760 hourly observations).

**Pollutants**: NO₂, PM₂.₅, PM₁₀, O₃.

MY1 is a well-known kerbside site with dense coverage of all regulatory
pollutants, chosen as a canonical comparison dataset because it exercises
every metric on realistic roadside concentrations.

## Loading the same raw input in both environments

The R reference — not executed here, shown for documentation:

```{r}
#| eval: false
library(openair)
raw <- importAURN(site = "my1", year = 2023)
```

The Python side reads the same raw data produced by `importAURN`, via
the CSV fixture:

```{python}
#| label: load-raw
import pandas as pd
from pathlib import Path

FIXTURES = Path("../../tests/fixtures/openair")
raw = pd.read_csv(FIXTURES / "aurn_my1_2023_raw.csv")
raw["date"] = pd.to_datetime(raw["date"], utc=True)
raw.head()
```

## Metric 1 — daily mean with 75% data-capture threshold

The standard LAQM annual-report metric: the mean of hourly values within
each calendar day, with the day dropped (NaN) if fewer than 75% of hours
are valid.

R openair:

```{r}
#| eval: false
daily <- timeAverage(raw, avg.time = "day", data.thresh = 75)
```

aeolus:

```{python}
#| label: aeolus-daily
from datetime import datetime, timezone
from aeolus.metrics import time_average

# Reshape openair's wide-format raw data into aeolus's long-format schema
pollutants = ["no2", "pm2.5", "pm10", "o3"]
long = raw.melt(
    id_vars=["date", "site", "code"],
    value_vars=pollutants,
    var_name="measurand",
    value_name="value",
)
long["measurand"] = long["measurand"].map(
    {"no2": "NO2", "pm2.5": "PM2.5", "pm10": "PM10", "o3": "O3"}
)
aeolus_input = pd.DataFrame({
    "site_code": long["code"],
    "date_time": long["date"],
    "measurand": long["measurand"],
    "value": pd.to_numeric(long["value"], errors="coerce"),
    "units": "ug/m3",
    "source_network": "AURN",
    "ratification": "None",
    "created_at": datetime(2024, 1, 1, tzinfo=timezone.utc),
})

aeolus_daily_long = time_average(aeolus_input, freq="D", data_thresh=0.75)
aeolus_daily = aeolus_daily_long.pivot_table(
    index="date_time", columns="measurand", values="value", aggfunc="first"
).reset_index().rename(columns={"date_time": "date"})
aeolus_daily = aeolus_daily.rename(
    columns={"NO2": "no2", "PM2.5": "pm2.5", "PM10": "pm10", "O3": "o3"}
)
```

### Comparison

```{python}
#| label: compare-daily
import numpy as np

openair_daily = pd.read_csv(FIXTURES / "aurn_my1_2023_daily_mean_thresh75.csv")
openair_daily["date"] = pd.to_datetime(openair_daily["date"], utc=True)

merged = aeolus_daily.merge(
    openair_daily, on="date", suffixes=("_aeolus", "_openair")
)

rows = []
for p in pollutants:
    a = merged[f"{p}_aeolus"].to_numpy()
    o = merged[f"{p}_openair"].to_numpy()
    both = ~np.isnan(a) & ~np.isnan(o)
    diffs = np.abs(a[both] - o[both])
    rows.append({
        "pollutant": p.upper(),
        "n_days_compared": int(both.sum()),
        "n_days_nan_both": int((np.isnan(a) & np.isnan(o)).sum()),
        "max_abs_diff": f"{diffs.max():.2e}" if len(diffs) else "—",
        "mean_abs_diff": f"{diffs.mean():.2e}" if len(diffs) else "—",
    })
pd.DataFrame(rows)
```

All four pollutants match to ~10⁻¹⁴ µg/m³ — floating-point noise level,
indicating bit-identical algorithms modulo the order of summation.

### Time-series overlay

```{python}
#| label: plot-overlay
#| fig-cap: "aeolus vs openair daily NO₂ — red dots visible only where values disagree."
import matplotlib.pyplot as plt

fig, ax = plt.subplots(figsize=(10, 4))
ax.plot(merged["date"], merged["no2_openair"], label="openair", lw=1.2)
ax.plot(merged["date"], merged["no2_aeolus"], label="aeolus", lw=1.2, linestyle="--")

disagreements = np.abs(merged["no2_aeolus"] - merged["no2_openair"]) > 1e-6
if disagreements.any():
    ax.scatter(
        merged.loc[disagreements, "date"],
        merged.loc[disagreements, "no2_aeolus"],
        color="red", s=30, zorder=5, label="disagreement (>1e-6)",
    )

ax.set_ylabel("NO₂ (µg/m³)")
ax.set_xlabel("Date")
ax.legend()
ax.grid(True, alpha=0.3)
plt.tight_layout()
```

::: {.callout-tip}
### Re-running the reference

If you want to verify the reference outputs yourself, install R and openair
(version pinned in the provenance table above), then run:

```bash
Rscript scripts/validation/generate_openair_fixtures.R
```

This regenerates every file in `tests/fixtures/openair/`. Compare the
output to what's checked in — it should be bit-identical.
:::

## Status

| Metric | Tolerance | Status |
|--------|-----------|--------|
| `time_average(freq='D', data_thresh=0.75)` | 10⁻¹² µg/m³ | ✅ Pass |

More metrics will be added to this comparison as aeolus's metrics module
grows. Each is gated by a CI test in `tests/test_openair_parity.py`; if
this page renders, every listed metric is passing.
